Analysis-ready VCF at Biobank scale using Zarr

### 2024-04-30 paper.tex

Zarr is flexible in allowing different compression codecs and 
pre-compression filters to be specified on a per-array basis.
Two key technologies used by Zarr are the Blosc
% This seems like the best citation for Blosc? It is mentioned here
meta-compressor~\cite{alted2010modern}
and Zstandard compression algorithm~\citep{collet2021rfc}.
Blosc is a high-performance compressor optimised for numerical
data which uses ``blocking''~\citep{alted2010modern} to 
optimise CPU-cache access patterns, as well as highly optimised
bit and byte shuffle filters.  Remarkably, on highly 
compressible datasets, Blosc decompression can be faster 
than \texttt{memcpy}.
Blosc is written in C, with APIs for C, Python, Julia, Rust
and others.
Blosc is a ``meta-compressor'' because it provides different 
access to several different compression codecs. The 
Zstandard compressor~\citep{collet2021rfc} is of particular 
interest here as it achieves very high compression ratios
with fast decompression speeds [see fig]. 
% This may seem a bit off topic, but want to reassure reader that 
% this isn't some niche technology
Zstandard is also used in several recent VCF compression 
methods~\citep[e.g.][]{lefaive2021sparse,wertenbroek2022xsi}.
% LZ4 is also notable - any citations for this?

This combination of Zarr's simple and cloud-friendly storage 
of data chunks with state-of-the-art compression methods has 
led to Zarr gaining significant traction in large-scale 
scientific computing, in particular in the geosciences.

Multi terabyte numerical weather prediction models~\citep{gowan2022using}

CMIP6 dataset in Google Cloud~\cite{cmip6_dataset}

The Open GeoSpatial consortium has formally recognised Zarr as a community
standard~\cite{ogc_zarr2_standard}
and has formed
a new GeoZarr Standards Working Group to establish a Zarr encoding for
geospatial data~\cite{ogc_geozarr_news}.

Zarr is being considered a potential format for migrating the 
Copernicus global land service to the cloud~\citep{abernathey2021opening}

ITS\_LIVE ice velocity dataset~\citep{fahnestock2023mappin} using 
the Zarr.js Javascript interface to visualise the ``massive''
dataset [how big is it?]

>>>>

Zarr is flexible in allowing different compression codecs and 
pre-compression filters to be specified on a per-array basis
(see section XXX).
Two key technologies used by Zarr are the Blosc
% This seems like the best citation for Blosc? It is mentioned here
meta-compressor~\cite{alted2010modern}
and Zstandard compression algorithm~\citep{collet2021rfc}.
Blosc is a high-performance compressor optimised for numerical
data which uses ``blocking''~\citep{alted2010modern} to 
optimise CPU-cache access patterns, as well as highly optimised
bit and byte shuffle filters.  Remarkably, on highly 
compressible datasets, Blosc decompression can be faster 
than \texttt{memcpy}.
Blosc is written in C, with APIs for C, Python, Julia, Rust
and others.
Blosc is a ``meta-compressor'' because it provides different 
access to several different compression codecs. The 
Zstandard compressor~\citep{collet2021rfc} is of particular 
interest here as it achieves very high compression ratios
with fast decompression speeds [see fig]. 
% This may seem a bit off topic, but want to reassure reader that 
% this isn't some niche technology
Zstandard is also used in several recent VCF compression 
methods~\citep[e.g.][]{lefaive2021sparse,wertenbroek2022xsi}.
% LZ4 is also notable - any citations for this?

Scientific datasets are increasingly overwhelming the classical
model of downloading and analysing locally, and are migrating to 
centralised cloud repositories \citep{abernathey2021cloud}.
The combination of Zarr's simple and cloud-friendly storage 
of data chunks with state-of-the-art compression methods has 
led to Zarr gaining significant traction in these settings.
Multiple petabyte-scale datasets are now stored using 
Zarr~\cite[e.g.][]{gowan2022using, % weather model data
fahnestock2023mappin, % ITS_LIVE dataset
cmip6_dataset}
or under active consideration for migration~\citep{abernathey2021opening}.
The Open GeoSpatial consortium has formally recognised Zarr as a community
standard~\cite{ogc_zarr2_standard}
and has formed
a new GeoZarr Standards Working Group to establish a Zarr encoding for
geospatial data~\cite{ogc_geozarr_news}.
